\section{Nikola module overiew reference} \label{section:nikola-reference}

In this section we provide a small overview of central modules and namespaces
in Nikola for reference. These modules all reside in the
\texttt{Data.Array.Nikola} namespace.

\subsection{Frontend modules}
\begin{description}

  \item[\texttt{Exp}] module. Exports the type \texttt{Exp t a} and typeclass
    \texttt{IfThenElse} for use with rebindable syntax. The type \texttt{Exp t
    a} will sometimes appear as qualified name \texttt{E.Exp} in case of
    ambiguity.

  \item[\texttt{Array}] module. Defines \texttt{IsArray r e} typeclass with
    associated type \texttt{Array r sh e}. It also defines the
    \texttt{Source r e} typeclass for arrays that support indexing.

  \item[\texttt{Shape}] module. Defines types \texttt{Z}, \texttt{sh :. a}
    and the typeclass \texttt{Shape sh}, with instances for \texttt{Z} and
    \texttt{sh :. a}.

  \item[\texttt{Repr}] namespace. Contains modules definining the array
    representations for Push, Delayed and Global arrays through
    \texttt{IsArray} instances.

  \item[\texttt{Backend.CUDA.Haskell}, \texttt{Backend.CUDA.TH}] modules.
    Define the \texttt{compile} function for runtime and compiletime usage
    respectively.

\end{description}

\subsection*{Backend modules}
\begin{description}

  \item[\texttt{Language.Syntax}] module. Exports the \texttt{Exp} datatype,
    which is a first-order abstract syntax for primitive Nikola expressions.
    The abstract syntax contains constructs for manipulating memory arrays,
    looping and annonymous functions. It also exports the datatypes
    \texttt{Type} and \texttt{ScalarType} used as church style type annotations
    for \texttt{Exp} values. In the case of ambiguity we quailfy \texttt{Exp}
    as \texttt{S.Exp}.

  \item[\texttt{Language.Check}] module. Contains a type checker and simple type
    inference for \texttt{Exp} values. Also defines the typeclass
    \texttt{MonadCheck m}, to allow type checking to be performed in any monad
    that defines how to access to the typing environment

  \item[\texttt{Language.Monad}] module. Defines the continuation passing
    \texttt{R r a} monad, and the type alias \texttt{type P a = R Exp a}.
    Also defines the combinators \texttt{shift} and \texttt{reset} for
    delimited continuations \cite{waddler1994monads}.

  \item[\texttt{Language.Reify}] module. Exports the \texttt{Reifiable a b}
    typeclass. It provides the operation \texttt{reify :: a -> R b b},
    which is the central piece in converting frontend data types
    \texttt{Exp t a} and \texttt{Array r sh e} to primitive \texttt{Exp}.

  \item[\texttt{Language.Optimize}] namespace and module. Provides various
    optimizations on Nikola \texttt{S.Exp} values exclusively. Most notable is
    the detection of let-sharing, see \cite{mainland2010nikola}. Marking
    parallel for-loops at the top level to be translated to CUDA kernels is
    also treated as an optimisation in Nikola.

  \item[\texttt{Backend.C.Codegen}] module. Generates C code from \texttt{S.Exp}
    values, represented as C abstract syntax.

\end{description}

Apart from these modules Nikola also provides various implementations of
vectors in the \texttt{Data.Vector.CUDA} namespace. These vectors are
stored in device memory and integrate with \texttt{compile} to allow
control of the timing of host-device data transfers.

Furthermore these vector implementations are integrated with Repa as additional
array representation types in modules in the \texttt{Data.Array.Repa.Repr.CUDA}
namespace.

\subsection{Array representation capabilities.}

The operations that arrays may be used with is modeled with the use of type
classes.

\begin{itemize}

  \item The \texttt{Source r e} type class specifies an \texttt{index}
    function that allow a consumer to extract values of type \texttt{e} from an
    array with representation \texttt{r} and contents \texttt{e}.

    Instances:
    \begin{itemize}
      \item \texttt{Source D e}
      \item \texttt{IsElem e => Source G e}
    \end{itemize}

  \item The \texttt{Target r e} type class enables mutable access of
    \texttt{r}-arrays in the \texttt{P} monad.

    Instances:
    \begin{itemize}
      \item \texttt{IsElem e => Target G e}
    \end{itemize}

  \item The \texttt{Load r sh e} type class enables manifestation of
    \texttt{r}-arrays into \texttt{Target r' e} arrays in the \texttt{P} monad.

    Instances:
    \begin{itemize}
      \item \texttt{Shape sh => Load sh D}
      \item \texttt{Shape sh => Load sh PSH}
    \end{itemize}
\end{itemize}
