\documentclass[10pt,a4paper,final,oneside,openany,article]{memoir}

\input{prelude.tex}

\title{Vector languages for financial algorithms \\
      \small{Master thesis project description}
}
\author{
  Martin Dybdal -- \texttt{dybber@dybber.dk} -- \texttt{dpr964} \\
  Philip Lykke Carlsen -- \texttt{plcplc@gmail.com} -- \texttt{jhg363}
\\
University of Copenhagen}

\date{\today}

\bibliography{../bibliography/bibliography}


\begin{document}
\maketitle

\chapter{Introduction}
This is the project description of our joint master thesis work. The
master thesis will be supervised by Ken Friis Larsen and Cosmin
Oaneca.

\chapter{Motivation}
% What is the overall subject matter and why is it interesting to do
% further research in this subject
Finance is the subfield of economics that has to do with valuing
assets, making investment decisions and measuring the risk incurred by
such decisions. The analyses used to make decisions and performing
risk management are complex and computationally intensive. What is
more, is that some of these problems have to be solved very
efficiently such that results are still relevant when the computation
is complete.

The current understanding of mathematics does not yet permit
closed-form analytical solutions for most of the problems that
financial institutions need to solve. Thus they turn to numerical
approximation algorithms where the Monte Carlo method is a popular
tool for its versatility and ease of parallelization.

To meet this end, the computational model employed by \textit{vector
  languages} also known as \textit{collection-oriented languages}
\cite{blelloch1990vector} appears to be a promising fit. Operations in
vector languages are naturally lifted to apply directly on vector
elements in bulk, and this allows for a parallel implementation that
fits well into modern SIMD computer architectures, such as graphics
processing units (GPUs).

The improvements in parallel vector languages and tools for general
purpose programming of GPUs will not only be applicable to financial
problems. High performance computing has a wide audience in the
scientific community, and improving the infrastructure for computation
will be useful in for example weather forecasting, bioinformatics and
simulation of experiments such as chemical reactions. A difference is
that the real-time requirements of certain financial applications are
not that important in these fields.

\chapter{Project goals}
% In short describe the experiment we are going to make
In this project we will first seek to find the extent to which
current parallel vector languages are applicable for the financial
domain, by implementing a set of predetermined pricing algorithms in
existing languages.

We will then determine how we can improve the state-of-the-art of
vector languages in the context of pricing algorithms. We will do so
in terms of expressiveness, safety and efficiency, by implementing
enhancements for one or more of the existing languages.

\chapter{Elaboration}
% Elaborate on the experiment, e.g. how it should be conducted and
% what others have done previously

%\todo{mention GPU programming and highly parallel computer architecture}

As the first task in our project, we will conduct a survey among
existing vector languages, such as
Accelerate\cite{chakravarty2011accelerating}, Nikola
\cite{mainland2010nikola}, Repa \cite{keller2010regular}, Feldspar
\cite{axelsson2010feldspar}, Obsidian \cite{svensson2011obsidian} and
\textsc{Nesl} \cite{nesl} (tentative list). These vector languages
take different approaches and have different limitations and
advantages.

The survey will consist of writing a set of realistic example programs
from the financial domain, specifically option pricing algorithms, and
comparing against handcrafted OpenCL code. Example algorithms includes
basic algorithms such as \textit{Longstaff and Schwartz},
\textit{Black-Scholes}, \textit{Binomial option pricing} and a real
world financial application (provided by the company LexiFi).

The ultimate goal of the project is to get an overview of existing
languages and libraries, and improve the most promising candidate
towards more efficient execution of our example applications on
GPUs. We intend the improvements to be general enough that they will
also benefit other applications. Some of the libraries mentioned above
do not target graphics processing units (\textit{Repa} and
\textit{Feldspar}), but are included to provide a more complete
picture of the current directions of vector languages.

The improvements fall in two categories, improvements in terms of
expressivity and in terms of execution efficiency. For example, the
language Accelerate goes a long way to ensure efficiency, but does so
by restricting the set of language primitives, which makes it
impossible to write many real world programs. If we were to improve
Accelerate it would thus primarily be in the direction of
expressivity. For other languages we might go in the other direction,
and focus on optimizations such as memory coalescing, stream fusion
and limiting the number of divergent branches.

Finally we will evaluate the improvements carried out and compare with
the surveyed vector languages.

\chapter{Learning objectives}
% What are we (as students) going to learn from performing this
% project
After completing this master thesis we will be able to:
\begin{itemize}
%\item Explain the financial applications of the pricing algorithms
%  mentioned above, and the basics of financial contracts.
\item Describe the different types of financial products (options) and
  how option pricing is performed, with different classes of methods.
\item Describe current vector languages and present relevant criteria
  for their comparison and empirical evaluation.
\item Perform empirical evaluation of vector languages according to
  the criteria developed above, by implementing various pricing
  methods.
\item Write programs for GPUs and apply optimisations.
\item Implement and evaluate extensions/enhancements (see above) of
  current vector languages.
\end{itemize}

\chapter{Scope}
% What are we NOT doing. That is, which limitations do we make
\begin{itemize}
\item The financial applications are only used as examples, and we
  will only superficially sketch their theory and industry usage.
%\item We will only look at example applications from the financial domain
\end{itemize}

%\newpage
\defbibheading{bibliography}{\chapter{References}} 
\printbibliography

%\printbibliography[keyword=dsls,heading=subbibliography,
%  title={Domain specific language infrastructure}]
\end{document}
